A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-19 tex/body.tex
as discussed on Skype

%\newcommand\DISC[1]{t \teq \const{C}_{#1}\bigl(\const s^1_{#1}( t ), \ldots,\const s^{n_{#1}}_{#1}( t )\bigr)}
\newcommand\DISC[1]{\const{d}_{#1}(t)}

@@

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Ec}{E}
%\newcommand{\Ec}{\mathcalx{E}}
%\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Fc}{F}
%\newcommand{\Fc}{\mathcalx{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\Terms(\Ec)}
\newcommand{\tcEc}{\Terms^\AST(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\ror}{\parallel}
%\newcommand{\ror}{{\&}}
%\newcommand{\ROR}{\kern.45em \ror \kern.45em}
\newcommand\ROR{\betweenantes}
\newcommand{\tpath}[2]{\mathcalx{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcalx{P}_{ #2 \rightarrow #3 }( #1 )}
%\newcommand{\interp}[2]{#1\llbracket #2 \rrbracket}
%%% @ANDY: I prefer a notation that puts more emphasis on #2 (which is, in essence,
%%% the object that is returned) and less on #1
%\newcommand{\interp}[2]{\llbracket #2 \rrbracket_{#1}}
%%% And one that doesn't look so semantic
%\newcommand{\interp}[2]{\lceil #2 \rceil_{#1}}
\newcommand{\interp}[2]{\langle \smash{#2} \rangle_{#1}}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\JJJJ}{\mathcal{\!J\!}}
\newcommand{\J}{\mathcalx{J}}
\newcommand{\M}{\mathcalx{M}}
\newcommand{\Val}{\mathcalx{A}\vvthinspace}
\newcommand{\ValC}{\mathcalx{A}^\star}
\newcommand{\Var}{\mathcalx{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\vvthinspace\widetilde{#1}\vvthinspace}
\newcommand{\VAREC}[1]{\widetilde{\vphantom{\scriptstyle x^i}\smash{#1}}}
%\newcommand{\nf}[1]{{{#1}{\downarrow}}}
\newcommand{\nf}[1]{\lfloor#1\rfloor}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
%\newcommand{\muvar}{\mathcalx{Var}}
\newcommand{\muvar}{\mathrm{Var}} %% cf. FV

@@

To simplify the presentation, we make a few assumptions about
$\Sig$.
First, all codatatypes in $\Codata$ are corecursive. This is not restrictive
because noncorecursive codatatypes can be seen as nonrecursive
datatypes.
Second, all types $\tau \in \Nondata$ have infinite cardinality.
This is not very limitative, because each interpreted type in $\Nondata$
having finite cardinality~$n$
can be replaced by an enumeration datatype with $n$~nullary constructors.
As for uninterpreted types, since the constraints in $\Ec$ are ground, they cannot entail an upper
bound on the cardinality of any such type, so it is safe to consider them infinite.

@@

\paragraph{Phase 3: Branching \rm(\figurename~\ref{fig:split-rule}).}
If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
the equivalence class of $t$ must contain a
$\delta$ constructor term.
This is enforced in the third phase by the \rn{Split} rule.
A companion rule, \rn{Single}, focuses on the degenerate case where two
terms %from $\tEc$
have a corecursive singleton type
%(Section~\ref{sec:the-theory-of-co-datatypes})
and must therefore be equal. Both
\rn{Split}'s finiteness assumption %on the datatype
and \rn{Single}'s singleton
constraint %on the codatatype
can be evaluated statically
based on a recursive computation of the cardinalities of the
constructors' argument types. This step relies on the assumption that
ordinary types are infinite.

@@

\begin{theorem}[Termination]%
\label{thm:t}%
\afterDot
All derivation trees are finite.
\end{theorem}
\begin{proof}
Consider a derivation tree with root node $\Ec$. Let $D \subseteq \tEc$ be the
set of terms whose types are finite datatypes, and let $\SSS \subseteq \tEc$ be
the set of terms occurring as arguments to selectors. For each term $t \in
D$\negvthinspace, let
\begin{conf}
$\smash{\SSSS 0_t} = \{ t \}$ and
$\smash{\SSSS {i+1}_t} = \smash{\SSSS i_t \mathrel\cup \{ \const s( u ) \mid u\typ{\delta} \in \SSSS i_t{,}}~\allowbreak\smash{ \delta \in \Data{,}~ \left|\delta\right| \text{ is finite}{,}~ \const{s} \in \Sel^{\delta}}  \}$,
\end{conf}%
\begin{rep}
\begin{align*}
\smash{\SSSS 0_t} & = \{ t \}
& \smash{\SSSS {i+1}_t} & = \smash{\SSSS i_t \mathrel\cup \{ \const s( u ) \mid u\typ{\delta} \in \SSSS i_t{,}~ \delta \in \Data{,}~ \left|\delta\right| \text{ is finite}{,}~ \const{s} \in \Sel^{\delta}}  \}
\end{align*}
\end{rep}%
%for $i \geq 0$, let ;
and let $\SSSS \AST_t$ be the limit of this sequence.
This is a finite set for each $t$ because there can only be finitely many
values of the same type as $t$.
%%% TODO: Clarify above step in report.
Let $\SSSS \AST$ be the union of all sets $\SSSS \AST_t$ where $t \in D$,
and let $\tcEc$ be the set of subterms of
%
$\Ec \mathrel\cup \{ \const C_{\negvthinspace j}\bigl( \const s^1_{\negvthinspace j}( t ), \ldots, \const s^{\smash{\,n_{\negvthinspace j}}}_{\negvthinspace j}( t ) \bigr) \mid t\typ{\delta} \in \SSS \mathrel\cup \SSSS \AST{,}~ \const C_{\negvthinspace j} \in \Ctr^\delta \}$.
%
In a derivation tree with root node~$\Ec$,
it can be shown by %case analysis
induction on the rules of the calculus that each non-root node $\Fc$ is such that
$\Terms(\Fc) \subseteq \tcEc$, and hence contains an equality between two terms from $\tcEc$ not occurring in its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\left| \smash{\tcEc}\vphantom{X} \right|{\!}^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

@@

However, the presence of quantified formulas poses an additional challenge in the context of (co)datatypes:
Since quantified formulas may entail cardinality constraints on uninterpreted types $\ty{u}$,
we may no longer assume that they are of infinite cardinality.
When assuming that a finite type has infinite cardinality, the calculus presented in this paper is \relax{incomplete}
since it may fail to recognize cases when the preconditions of \rn{Split} and \rn{Single} hold.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete when quantified formulas are present.
Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, consider the $\ty{stream}_{\,\ty{u}}$ codatatype, where $\ty{u}$ is uninterpreted.
This type has cardinality precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F$.
\begin{rep}In other words, either the cardinality of $\ty{u}$ is greater than one or $s$ is equal to $t$.\end{rep}

%\paragraph{Optimizations.}
The implementation of the decision procedure uses several optimizations
following the lines of Barrett et al. \cite{barrett-et-al-2007}.
%and are not reflected in the presentation of the calculus.
We briefly mention the main ones.
%
Discriminators are part of the signature instead
of being abbreviations.
We have found that this leads to better performance since it introduces terms less eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several
rules \cite{barrett-et-al-2007},
which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
If $\const s^k_{\negvthinspace j}( t ) \in \tEc$ and $t =
\const{C}_{\negvthinspace j}( u_1, \ldots, u_n )$, we immediately add
$\smash{\const s^k_{\negvthinspace j}( t ) \teq u_k}$ to $\Ec$, whereas the presented calculus
would apply \rn{Split} and \rn{Inject} before adding this equality. 
\begin{rep}  % lower priority to mention this
To reduce the number of unique constraints considered by the calculus, we
compute a normal form for constraints as a preprocessing step. In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term
ordering, replace $\const{C}_{\negvthinspace j}( \vec t ) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$ by $t_k$, and replace
occurrences of discriminators $\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t ) )$ by $\top$ or $\bot$
based on whether $j = \jPrime$.
\end{rep}

>>>>

\newcommand\DISC[1]{t \teq \const{C}_{#1}\bigl(\const s^1_{#1}( t ), \ldots,\const s^{n_{#1}}_{#1}( t )\bigr)}
%\newcommand\DISC[1]{\const{d}_{#1}(t)}

@@

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Ec}{E}
%\newcommand{\Ec}{\mathcalx{E}}
%\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Fc}{F}
%\newcommand{\Fc}{\mathcalx{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\Terms(\Ec)}
\newcommand{\tcEc}{\Terms^\AST(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\ror}{\parallel}
%\newcommand{\ror}{{\&}}
%\newcommand{\ROR}{\kern.45em \ror \kern.45em}
\newcommand\ROR{\betweenantes}
\newcommand{\tpath}[2]{\mathcalx{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcalx{P}_{ #2 \rightarrow #3 }( #1 )}
%\newcommand{\interp}[2]{#1\llbracket #2 \rrbracket}
%%% @ANDY: I prefer a notation that puts more emphasis on #2 (which is, in essence,
%%% the object that is returned) and less on #1
%\newcommand{\interp}[2]{\llbracket #2 \rrbracket_{#1}}
%%% And one that doesn't look so semantic
%\newcommand{\interp}[2]{\lceil #2 \rceil_{#1}}
\newcommand{\interp}[2]{\langle \smash{#2} \rangle_{#1}}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\JJJJ}{\mathcal{\!J\!}}
\newcommand{\J}{\mathcalx{J}}
%\newcommand{\M}{\mathcalx{M}}
\newcommand{\M}{\mathcalx{J}}
\newcommand{\Val}{\mathcalx{A}\vvthinspace}
\newcommand{\ValC}{\mathcalx{A}^\star}
\newcommand{\Var}{\mathcalx{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\vvthinspace\widetilde{#1}\vvthinspace}
\newcommand{\VAREC}[1]{\widetilde{\vphantom{\scriptstyle x^i}\smash{#1}}}
%\newcommand{\nf}[1]{{{#1}{\downarrow}}}
\newcommand{\nf}[1]{\lfloor#1\rfloor}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
%\newcommand{\muvar}{\mathcalx{Var}}
\newcommand{\muvar}{\mathrm{Var}} %% cf. FV

@@

To simplify the presentation, we make a few assumptions about
$\Sig$.
First, all codatatypes in $\Codata$ are corecursive. This is reasonable
because noncorecursive codatatypes can be seen as nonrecursive
datatypes.
Second, all types $\tau \in \Nondata$ have infinite cardinality.
This is not very restrictive, because in principle each interpreted type in
$\Nondata$ having finite cardinality~$n$
can be replaced by a %n enumeration
datatype with $n$~nullary constructors
\cite{barrett-et-al-2007}.
As for uninterpreted types, since the constraints in $\Ec$
are ground, they cannot entail an upper
bound on the cardinality of any such type, so it is safe to consider them infinite.

@@

\paragraph{Phase 3: Branching \rm(\figurename~\ref{fig:split-rule}).}
If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
the equivalence class of $t$ must contain a
$\delta$ constructor term.
This is enforced in the third phase by the \rn{Split} rule.
A companion rule, \rn{Single}, focuses on the degenerate case where two
terms %from $\tEc$
have a corecursive singleton type
%(Section~\ref{sec:the-theory-of-co-datatypes})
and must therefore be equal. Both
\rn{Split}'s finiteness assumption %on the datatype
and \rn{Single}'s singleton
constraint %on the codatatype
can be evaluated statically
based on a recursive computation of the cardinalities of the
constructors' argument types.
%This step relies on the assumption that ordinary types are infinite.

@@

\begin{theorem}[Termination]%
\label{thm:t}%
\afterDot
All derivation trees are finite.
\end{theorem}
\begin{proof}
Consider a derivation tree with root node $\Ec$. Let $D \subseteq \tEc$ be the
set of terms whose types are finite datatypes, and let $\SSS \subseteq \tEc$ be
the set of terms occurring as arguments to selectors. For each term $t \in
D$\negvthinspace, let
\begin{conf}
$\smash{\SSSS 0_t} = \{ t \}$ and
$\smash{\SSSS {i+1}_t} = \smash{\SSSS i_t \mathrel\cup \{ \const s( u ) \mid u\typ{\delta} \in \SSSS i_t{,}}~\allowbreak\smash{ \delta \in \Data{,}~ \left|\delta\right| \text{ is finite}{,}~ \const{s} \in \Sel^{\delta}}  \}$,
\end{conf}%
\begin{rep}
\begin{align*}
\smash{\SSSS 0_t} & = \{ t \}
& \smash{\SSSS {i+1}_t} & = \smash{\SSSS i_t \mathrel\cup \{ \const s( u ) \mid u\typ{\delta} \in \SSSS i_t{,}~ \delta \in \Data{,}~ \left|\delta\right| \text{ is finite}{,}~ \const{s} \in \Sel^{\delta}}  \}
\end{align*}
\end{rep}%
%for $i \geq 0$, let ;
and let $\SSSS \AST_t$ be the limit of this sequence.
This is a finite set for each $t$, because in the absence of (co)recursion
one eventually reaches a point where no selector can be applied.
%%% TODO: Clarify above step in report.
Let $\SSSS \AST$ be the union of all sets $\SSSS \AST_t$ where $t \in D$,
and let $\tcEc$ be the set of subterms of
%
$\Ec \mathrel\cup \{ \const C_{\negvthinspace j}\bigl( \const s^1_{\negvthinspace j}( t ), \ldots, \const s^{\smash{\,n_{\negvthinspace j}}}_{\negvthinspace j}( t ) \bigr) \mid t\typ{\delta} \in \SSS \mathrel\cup \SSSS \AST{,}~ \const C_{\negvthinspace j} \in \Ctr^\delta \}$.
%
In a derivation tree with root node~$\Ec$,
it can be shown by %case analysis
induction on the rules of the calculus that each non-root node $\Fc$ is such that
$\Terms(\Fc) \subseteq \tcEc$, and hence contains an equality between two terms from $\tcEc$ not occurring in its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\left| \smash{\tcEc}\vphantom{X} \right|{\!}^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

@@

However, the presence of quantified formulas poses an additional challenge in the context of (co)datatypes:
Since quantified formulas may entail cardinality constraints on uninterpreted types $\ty{u}$,
we may no longer assume that they are of infinite cardinality.
When assuming that a finite type has infinite cardinality, the calculus
presented in this paper is \relax{incomplete}
since it may fail to recognize cases where \rn{Split} and \rn{Single} should be
applied.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete when quantified formulas are present.
Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, consider the $\ty{stream}_{\,\ty{u}}$ codatatype, where $\ty{u}$ is uninterpreted.
This type has cardinality precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F$.
\begin{rep}In other words, either the cardinality of $\ty{u}$ is greater than one or $s$ is equal to $t$.\end{rep}

%\paragraph{Optimizations.}
The implementation of the decision procedure uses several optimizations
following the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
%and are not reflected in the presentation of the calculus.
We briefly mention the main ones.
%
Discriminators are part of the signature instead
of being abbreviations.
We have found that this leads to better performance since it introduces terms less eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several
rules \cite{barrett-et-al-2007},
which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
If $\const s^k_{\negvthinspace j}( t ) \in \tEc$ and $t =
\const{C}_{\negvthinspace j}( u_1, \ldots, u_n )$, we immediately add
$\smash{\const s^k_{\negvthinspace j}( t ) \teq u_k}$ to $\Ec$, whereas the presented calculus
would apply \rn{Split} and \rn{Inject} before adding this equality. 
\begin{rep}  % lower priority to mention this
To reduce the number of unique constraints considered by the calculus, we
compute a normal form for constraints as a preprocessing step. In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term
ordering, replace $\const{C}_{\negvthinspace j}( \vec t ) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$ by $t_k$, and replace
occurrences of discriminators $\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t ) )$ by $\top$ or $\bot$
based on whether $j = \jPrime$.
\end{rep}
